### 2020-07-27 Catuscia - final/01-introduction.tex
da dropbox

Natural Computing is an emerging area of research which has two main 
aspects: human designed computing inspired by nature, and computation 
performed in nature. Reaction Systems (RSs)~\cite{BEMR11} are 
% model of
% formal computation 
a rewriting formalism
%model 
inspired by the way biochemical reactions take place in living 
cells.  
This theory has already shown to be relevant in several different 
fields, such as computer science~\cite{MPR15}, 
biology~\cite{ABP14,CMMBM12,Az17,BarbutiGLM16}, 
molecular chemistry~\cite{OY16}.
%,
% and it
%is continuously developing, with a growing number of workshops and
%international meetings. 
Reaction Systems formalise the mechanisms of biochemical systems, 
such as {\em facilitation} and {\em inhibition}. 
As a qualitative approximation of the real biochemical reactions, they
% do not take
%into account the quantity of reagents effectively present. They
consider if a necessary reagent is or not present, and likewise they
consider if an inhibiting molecule is or not present. 
The  possible reactants and inhibitors are called `entities'.
RSs model in a direct way the interaction of a living cell
with the environment (called `context'). However, two RSs are seen
as independent models and do not interact.

%Formally, Reaction Systems are defined as a rewriting formalism.
In this paper, we present an encoding from RSs, 
to the open multiparty process algebra c\CNA,\footnote{After `chained Core Network 
Algebra'.} a variant of the {\tt link}-calculus~\cite{BodeiBB12,BBB17} without name mobility.
This formalism allows several processes to synchronise and 
communicate altogether, at the same time, with a new interaction
mechanism based on links and link chains. 
The initial motivation for introducing this mechanism was to encode
Mobile Ambients~\cite{CardelliG00}, getting a much stronger operational
correspondence than any available in the literature, such as the one in~\cite{B16}.
Later it was shown that the {\tt link}-calculus allowed one to easily encode calculi for biology equipped with membranes, as in~\cite{BodeiBBC14}.

We illustrate our embedding by means of some examples coming
both from the computer science and the biological field.
%and then we consider a more complex example, by modeling a RS 
%representing a regulatory network for \emph{lac} operon, presented in~\cite{CMMBM12}. 
We also show that our embedding preserves the main
features of RSs, and prove its correctness and completeness
from the operational semantics viewpoint.

Then, we present a methodology for verifying formally
properties of Reaction Systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
when one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to consider for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the Reaction Systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant introduced here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

Our main contributions are as follows:
\begin{itemize}
\item the behaviour of the context, for each single 
entity,
%\footnote{Entities that belong to the `domain' of a RS.}
can be
specified in a recursive way, as ordinary processes;
\item along the same lines, non-deterministic contexts can be seamlessly expressed in our setting; 
\item when deterministic contexts are considered, the c\CNA \
computation is deterministic and mimics the evolution of the corresponding RS;
\item we define an assertion language to specify local properties of 
Reaction Systems and exploit it to define a novel notion of behavioural equivalence for Reaction Systems, called bio-similarity;
\item we show that our assertions can be used to extend the Hennessy-Milner logic to our framework, in such a way that logical equivalence of processes coincide with bio-similarity;
\item we sketch how one can express the behaviour of entity mutation,
%change the behaviour of  each single entity $s$  in a
%way that a mutation $s'$ can be expressed; 
in such a way that the mutated entity
$s'$ can  take part to only a subset of rules requiring entity $s$;
\item we show that with a little coding effort, our encoding allows two RSs to
communicate; i.e. we can model scenarios where a subset of those entities that the context  can
provide, are instead provided by a second RS.
\end{itemize}

The main drawback of our proposal, is that the c\CNA \ translation is
verbose. Nevertheless it is clear that our
translation can be automatised by means of a proper front-end in
an implementation of the {\tt link}-calculus. 
The examples that we propose also show that some optimisations are possible
to reduce the coding when suitable assumptions are made about the provision of entities.

@@

\paragraph{Related work}

Process calculi have been used successfully to model
biological processes, see~\cite{BBDFH18} for a recent survey.
We are not aware of any SOS operational semantics for the RS; it seems not the case
that a deterministic behaviour, as the one of the RS, once the initial state is 
fixed, could be defined by inference rules that classically are applied 
to define non-deterministic transition systems.
The reversible computation paradigm for RS extends the RS framework  by
allowing  backward computations as well as forward computations; for this 
goal a set of inference rules for rewriting logic has been 
defined in~\cite{10.1007/978-3-319-73359-3_3}
to exactly keep trace of those elements that  dissolve in the next 
computation step.

A preliminary version of this paper appeared in~\cite{BBF19}.
There are several major differences w.r.t. the paper~\cite{BBF19}, which is here extended as follows:
\begin{itemize}
\item we present several new examples to illustrate our framework,
and give more detailed explanations about its use;
\item we introduce an assertion language to specify the properties of Reaction Systems;
\item we define the notion of bio-simulation to relate Reaction Systems and compare their behaviours;
\item we show that our assertion language allows to specify properties extending to our
framework the Hennessy-Milner logic;
\item we include here all proofs of main results.
\end{itemize}

\paragraph{Structure of the paper}

Section \ref{sec:rs} 
describes RSs and their semantics (interactive processes).
Section \ref{sec:ccna} describes briefly the c\CNA \ process algebra and
its operational semantics.
Section \ref{sec:trans} defines the embedding of RSs in
c\CNA \ processes and shows some simple examples to illustrate 
it.
Section~\ref{ex:examples} presents a couple of examples with computer science and biological applications.
%Section \ref{ex:lactose} shows a more complex example taken from
%the literature on RSs and illustrating a \emph{lac} operon.
Section~\ref{sec:biosimulation} presents a methodology for the formal verification of
properties of Reaction Systems that are expressed in a novel assertion language.
Section~\ref{sec:discussion} presents some features and 
advantages of our embedding for the compositionality of RSs.
Finally, Section~\ref{sec:conclusion}
discusses future work, and concludes.

>>>>

Natural Computing is an emerging area of research which has two main 
aspects: human designed computing inspired by nature, and computation 
performed in nature. Reaction Systems (RSs)~\cite{BEMR11} are a 
% model of
% formal computation 
rewriting formalism
%model 
inspired by the way biochemical reactions take place in living 
cells.  
This theory has already shown to be relevant in several different 
fields, such as computer science~\cite{MPR15}, 
biology~\cite{ABP14,CMMBM12,Az17,BarbutiGLM16}, and
molecular chemistry~\cite{OY16}.
%,
% and it
%is continuously developing, with a growing number of workshops and
%international meetings. 
Reaction Systems formalise the mechanisms of biochemical systems, 
such as {\em facilitation} and {\em inhibition}. 
As a qualitative approximation of the real biochemical reactions, they
% do not take
%into account the quantity of reagents effectively present. They
consider if a necessary reagent is or not present, and likewise they
consider if an inhibiting molecule is or not present. 
The  possible reactants and inhibitors are called `entities'.
RSs model in a direct way the interaction of a living cell
with the environment (called `context'). However, two RSs are seen
as independent models and do not interact.

%Formally, Reaction Systems are defined as a rewriting formalism.
In this paper, we present an encoding from RSs, 
to the open multiparty process algebra c\CNA\footnote{After `chained Core Network 
Algebra'.}, a variant of the {\tt link}-calculus~\cite{BodeiBB12,BBB17} without name mobility.
This formalism allows several processes to synchronise and 
communicate altogether, at the same time, with a new communicating
mechanism based on links and link chains. 
Our initial motivation for introducing this mechanism was to encode
Mobile Ambients~\cite{CardelliG00}, obtaining a much stronger operational
correspondence than any one available in the literature, such as the one in~\cite{B16}.
 This allowed us to easily encode calculi for biology equipped with membranes, as in~\cite{BodeiBBC14}.
Process calculi have been used successfully to model
biological processes, see~\cite{BBDFH18} for a recent survey.
We illustrate our embedding by means of some simple basic examples,
and then we consider a more complex example, by modeling a RS 
representing a regulatory network for \emph{lac} operon, presented in~\cite{CMMBM12}. We also show that our embedding preserves the main
features of RSs, and prove its correctness and completeness.
Our main contributions are as follows:
\begin{itemize}
\item the behaviour of the context, for each single 
entity,
%\footnote{Entities that belong to the `domain' of a RS.}
can be
specified in a recursive way as an ordinary process;
\item 
%\begin{color}{red}
our translation results in  a  c\CNA \ system where from
each state only one transition can be generated, thus the c\CNA \
computation is fully deterministic as that of the encoded RS;
%\end{color}
\item we can express the behaviour of entity mutation,
%change the behaviour of  each single entity $s$  in a
%way that a mutation $s'$ can be expressed; 
in such a way that the mutated entity
$s'$ can  take part to only a subset of rules requiring entity $s$;
\item with a little coding effort, two RSs can
communicate; i.e.  a subset of those entities that the context  can
provide, are then provided by a second RS.
\end{itemize}

The main drawback of our proposal, is that the c\CNA \ translation is
verbose. Nevertheless it is clear that our
translation can be automatised by means of a proper front-end in
an implementation of the {\tt link}-calculus.

@@

\paragraph{Structure of the paper.} Section \ref{sec:rs} 
describes RSs and their semantics (interactive processes).
Section \ref{sec:ccna} describes briefly the c\CNA \ process algebra and
its operational semantics.
Section \ref{sec:trans} defines the embedding of RSs in
c\CNA \ processes and shows some simple examples to illustrate 
it.
Section \ref{ex:lactose} shows a more complex example taken from
the literature on RSs and illustrates a \emph{lac} operon.
Section~\ref{sec:discussion} presents some features and 
advantages of our embedding for the compositionality of RSs.
Finally, Section~\ref{sec:conclusion}
discusses future work and concludes.

### 2020-07-27 Catuscia - final/02-ReactionSystem.tex
da dropbox

Natural Computing is concerned with human-designed computing inspired by 
nature as well as with computation taking place in nature.
The theory of Reaction Systems~\cite{BEMR11} 
was born in the field of Natural Computing 
to model the behaviour of biochemical reactions taking place in living cells. 
Despite its initial aim, this formalism has shown to be quite useful 
not only for modeling biological phenomena, but also for
the contributions which is giving to computer science~\cite{MPR15}, 
theory of computing, 
mathematics, biology~\cite{ABP14,CMMBM12,Az17,BarbutiGLM16}, 
and molecular chemistry~\cite{OY16}.
%
Here we briefly review the basic notions of RSs, see~\cite{BEMR11} for more details.

@@

The sets $R, I, P$ are also written $R_a, I_a, P_a$ and called the 
\emph{reactant set} of $a$, the  
\emph{inhibitor set} of $a$, and the \emph{product set} of $a$, respectively. 
All reactants are needed for the reaction to take place.
Any inhibitor blocks the reaction if it is present. Products are the outcome of the reaction.
Also,  $R_a \cup I_a$ is the set of the resources of $a$ and $ rac(S)$
denotes the set of all reactions in $S$.
%
Because $R$ and $I$ are non empty, all products are produced from at least one reactant and every reaction can be inhibited in some way. 
Sometimes artificial inhibitors are used that are never produced by any reaction.
For the sake of simplicity, in some examples, we will allow $I$ to be empty.
%Note that since $R_a$ and $I_a$ are both nonempty, we always 
%have $|M_a| \geq 2$.

@@

The set $S$ is called the \emph{background set} of ${\cal A}$, its elements are called \emph{entities},
they represent molecular substances (e.g., atoms, ions, molecules) that may be present
in the states of a biochemical system.
% modeled by $A$. 
The set $A$ is the set of \emph{reactions} of $\mathcal{A}$.
Since $S$ is finite, so is $A$: 
we denote by $|A|$ the number of reactions in $A$.

\begin{definition}[Reaction Result]
Let $W$ be a finite subset of $S$.
\begin{enumerate}
\item Let $a$ be a reaction in $S$. Then $a$ is enabled by $W$, denoted by $\mathit{en}_a(W)$, 
if $R_a \subseteq W$
and $I_a \cap W = \emptyset$. The result of $a$ on $W$, denoted by $\mathit{res}_a(W)$, 
is defined by:
$\mathit{res}_a(W) =P_a$, if $\mathit{en}_a(W)$, and $\mathit{res}_a(W) = \emptyset$ otherwise.
\item Let $A$ be a finite set of reactions. The result of $A$ on $W$, denoted 
by $\mathit{res}_A(W)$,  is defined
by: $\mathit{res}_A(W) = \bigcup_{a \in A} \mathit{res}_a(W)$.
\end{enumerate}
\end{definition}

%\medskip
%\noindent
%{\bf Important Assumptions}\\
The theory of Reaction Systems is based on the following assumptions.
\begin{itemize}
\item {\bf No permanency.} An entity of a set $W$ vanishes unless it is sustained 
by a reaction. This reflects the fact that a living cell would die for lack 
of energy, without chemical reactions.
\item {\bf No counting.} The basic model of RSs is very abstract and 
qualitative, i.e. the quantity of entities that are present 
in a cell is not taken into account. 
%It is just analyzed the behavior of 
%the system based on the reactions that take place.
\item {\bf Threshold nature of resources.} From the 
previous item,
we assume that either an entity is available and there is enough of it 
(i.e. there are no conflicts), or it is not available at all.
\end{itemize}

@@

\begin{definition}[Interactive Process]
Let ${\cal A} = (S, A)$ be a RS and let $n \geq 0$. 
An $n$-step
\emph{interactive process} in ${\cal A}$ is a pair $\pi = (\gamma, \delta)$ of 
finite sequences s.t.
%
%\begin{center}
$ \gamma=\{C_i\}_{i\in[0,n]}$  and $\delta=\{D_i\}_{i\in[0,n]} $
%$\gamma = C_0, \dots , C_n$ and $\delta = D_0, \dots , D_n$, 
%\end{center}
%\noindent
where 
$ C_{i}, D_{i}  \subseteq S$ for any $i\in[0,n]$, $D_0 = \emptyset$, and 
%$C_0,\dots, C_n, D_0,\dots, D_n  \subseteq S$, $D_0 = \emptyset$, and 
$D_i =  \mathit{res}_{A}(D_{i-1} \cup C_{i-1})$ for any $i \in [1,n]$.
% $i \in \{1,\dots, n\}$.
\end{definition}

Living cells are seen as open systems that continuously react with 
the external environment, in discrete steps. 
The sequence $\gamma$ is the {\em context sequence} of $\pi$ and 
represents the influence of the environment on the Reaction System.
The sequence $\delta$ is
the {\em result sequence} of $\pi$ and it is entirely determined by $\gamma$ and $A$.
%
The sequence $\tau = W_{0}, \ldots, W_{n}$ with $W_{i} = C_{i} \cup D_{i}$, 
for any $i \in [0, n]$
%, and $D_{0} = \emptyset$ 
is called a \emph{state sequence}.
%
Each state $W_{i}$ in a state sequence
is the union of two sets: the context $C_{i}$
at step $i$ and the result $D_i=\mathit{res}_{A}(W_{i-1})$ from the previous step.

@@

For technical reasons, we extend the notion of an interactive process to deal with 
infinite sequences.

\begin{definition}[extended interactive process]
Let ${\cal A}=(S,A)$ be a RS, and let $\pi =(\gamma,\delta)$  be an  $n$-step interactive process,
with $ \gamma=\{C_i\}_{i\in[0,n]}$ and  $\delta=\{D_i\}_{i\in[0,n]}$
%, $\delta= D_0, \dots D_n$.
Then, we let $\pi^\infty= (\gamma^\infty,\delta^\infty)$ be the extended interactive process of  $\pi$,
defined as $  \gamma^\infty=\{C_i'\}_{i\in\mathbb{N}}$, $\delta^\infty=\{D_i'\}_{i\in\mathbb{N}}$, where:
$$
C'_j = \left\{
\begin{array}{ll}
C_j & \mbox{if $j\in[0,n]$}\\
\emptyset & \mbox{otherwise ($j>n$)}
\end{array}
\right.
\qquad
D'_j = \left\{
\begin{array}{ll}
D_0 & \mbox{if $j=0$}\\
\mathit{res}_A(D_{j-1}' \cup C_{j-1}') & \mbox{if ($j\geq 1$)}
\end{array}
\right.
$$
% $C_j' = C_j$ for $j\in[0,n]$ and $C_j' =\emptyset$ for $j>n$,   $D_0' = D_0$ and  $D_j' = \mathit{res}_A(D_{j-1}' \cup C_{j-1}')$ for $j\geq 1$.
\end{definition}

Given an extended interactive process $\pi =(\gamma,\delta)$, we denote by $\pi^k$ the shift of $\pi$ starting at the $k$-th state sequence; formally we let $\pi^k=(\gamma^k,\delta^k)$ with
$\gamma^k=\{C_i'\}_{i\in\mathbb{N}}$, $\delta^k=\{D_i'\}_{i\in\mathbb{N}}$ with 
  $C_0'=C_k \cup D_k$, $D'_0 = \emptyset$, and   $C_i'=C_{i+k}$, $D_i'= D_{i+k}$ for any $i\geq 1$.

>>>>

Natural Computing is concerned with human-designed computing inspired by 
nature as well as with computation taking place in nature.
The theory of Reaction Systems~\cite{BEMR11} 
was born in the field of Natural Computing 
to model the behaviour of biochemical reactions taking place in living cells. 
Despite its initial aim, this formalism has shown to be quite useful 
not only for modeling biological phenomena, but also for
the contributions given to computer science~\cite{MPR15}, 
theory of computing, 
mathematics, biology~\cite{ABP14,CMMBM12,Az17,BarbutiGLM16}, 
and molecular chemistry~\cite{OY16}.
%
Here we briefly review the basic notions of RSs, see~\cite{BEMR11} for more details.

@@

The sets $R, I, P$ are also written $R_a, I_a, P_a$ and called the 
\emph{reactant set} of $a$, the  
\emph{inhibitor set} of $a$, and the \emph{product set} of $a$, respectively. 
All reactants are needed for the reaction to take place.
Any inhibitor blocks the reaction if it is present. Products are the outcome of the reaction.
Also,  $R_a \cup I_a$ is the set of the resources of $a$ and $ rac(S)$
denotes the set of all reactions in $S$.
%
Because $R$ and $I$ are non empty, all products are produced from at least one reactant and every reaction can be inhibited in some way. 
Sometimes artificial inhibitors are used that are never produced by any reaction.
For the sake of simplicity,  and without loss of generality, in some examples, we will allow $I$ to be empty.
%Note that since $R_a$ and $I_a$ are both nonempty, we always 
%have $|M_a| \geq 2$.

@@

The set $S$ is called the \emph{background set} of ${\cal A}$, its elements are called \emph{entities},
and they represent molecular substances (e.g., atoms, ions, molecules) that may be present
in the states of a biochemical system.
% modeled by $A$. 
The set $A$ is the set of \emph{reactions} of ${\cal A}$.
Since $S$ is finite, so is $A$: 
we denote by $|A|$ the number of reactions in $A$.

%\begin{color}{red}
\begin{definition}[Reaction Result]
Given a finite set of entities $S$, let $T\subseteq S$.
\begin{enumerate}
\item Let $a\in rac(S)$ be a reaction. Then $a$ is enabled by $T$, denoted by $en_a(T)$, 
if $R_a \subseteq T$
and $I_a \cap T = \emptyset$. 
\item Let $a\in rac(S)$ be a reaction. The result of $a$ on $T$, denoted by $res_a(T)$, 
is defined by:
$res_a(T) =P_a$ if $en_a(T)$, and $res_a(T) = \emptyset$ otherwise.
\item Let $A\subseteq rac(S)$ be a finite set of reactions. The result of $A$ on $T$, denoted 
by $res_A(T)$,  is defined
by: $res_A(T) = \bigcup_{a \in A} res_a(T)$.
\end{enumerate}
\end{definition}
%\end{color}

%\medskip
%\noindent
%{\bf Important Assumptions}\\
The theory of Reaction Systems is based on the following assumptions.
\begin{itemize}
\item {\bf No permanency.} An entity of a set $T$ vanishes unless it is sustained 
by a reaction. This reflects the fact that a living cell would die for lack 
of energy, without chemical reactions.
\item {\bf No counting.} The basic model of RSs is very abstract and 
qualitative, i.e. the quantity of entities that are present 
in a cell is not taken into account. 
%It is just analyzed the behavior of 
%the system based on the reactions that take place.
\item {\bf Threshold nature of resources.} From the 
previous item,
we assume that either an entity is available and there is enough of it 
(i.e. there are no conflicts), or it is not available at all.
\end{itemize}

@@

\begin{definition}[Interactive Process]
Let ${\cal A} = (S, A)$ be a RS and let $n \geq 0$. 
An $n$-step
\emph{interactive process} in ${\cal A}$ is a pair $\pi = (\gamma, \delta)$ of 
finite sequences s.t.
%
%\begin{center}
$ \gamma=\{C_i\}_{i\in[0,n]}$  and $\delta=\{D_i\}_{i\in[0,n]} $
%$\gamma = C_0, \dots , C_n$ and $\delta = D_0, \dots , D_n$, 
%\end{center}
%\noindent
where 
$ C_{i}, D_{i}  \subseteq S$ for any $i\in[0,n]$, $D_0 = \emptyset$, and 
%$C_0,\dots, C_n, D_0,\dots, D_n  \subseteq S$, $D_0 = \emptyset$, and 
$D_i =  res_{\cal A}(D_{i-1} \cup C_{i-1})$ for any $i \in [1,n]$.
% $i \in \{1,\dots, n\}$.
\end{definition}

Living cells are seen as open systems that continuously react with 
the external environment, in discrete steps. 
The sequence $\gamma$ is the {\em context sequence} of $\pi$ and 
represents the influence of the environment on the Reaction System.
The sequence $\delta$ is
the {\em result sequence} of $\pi$ and it is entirely determined by $\gamma$ and $A$.
%
The sequence $\tau = W_{0}, \ldots, W_{n}$ with $W_{i} = C_{i} \cup D_{i}$, 
for any $i \in [0, n]$
%, and $D_{0} = \emptyset$ 
is called a \emph{state sequence}.
%
Each state $W_{i}$ in a state sequence
is the union of two sets: the context $C_{i}$
at step $i$ and the result of the previous step.

@@

For technical reasons, we extend in a straightforward way the notion of an interactive process to deal with 
infinite sequences.

\begin{definition}[extended interactive process]
Let ${\cal A}=(S,A)$ be a RS, and let $\pi =(\gamma,\delta)$  be an  $n$-step interactive process,
with $ \gamma=\{C_i\}_{i\in[0,n]}$ and  $\delta=\{D_i\}_{i\in[0,n]}$.
%, $\delta= D_0, \dots D_n$.
Then, let $\pi'= (\gamma',\delta')$ be the extended interactive process of  $\pi =(\gamma,\delta)$,
defined as $  \gamma'=\{C_i'\}_{i\in\mathbb{N}}$, $\delta'=\{D_i'\}_{i\in\mathbb{N}}$, where
 $C_j' = C_j$ for $j\in[0,n]$ and $C_j' =\emptyset$ for $j>n$,   $D_0' = D_0$ and  $D_j' = res_A(D_{j-1}' \cup C_{j-1}')$ for $j\geq 1$.
\end{definition}

%\begin{color}{red}
The next example shows that the functioning of the interacting processes is deterministic, once the context is fixed.

\begin{example}
Let  ${\cal A}=(S,A)$ be a RS with  $S=\{s_1,s_2,s_3,s_4\}$, and  $A = \{ a_1,a_2\}$, where $a_1 = (\{s_1\},\{s_3\},\{s_1\})$,  $a_2 = (\{s_2\},\{s_4\},\{s_2\})$. 
 %It is easy to see that $r_1$ is incompatible with $r_2$ and with $r_3$, as $r_1$ requires the presence of $s_1$, instead $r_2$ and $r_3$ requires the absence of $s_1$. \\
% Now, reaction $r_1$ is only applied  when the entity configuration is $\{s_1,s_2\}$, or $\{s_1\}$.
 %If the configuration is $\{s_3,s_2\}$, then both $r_2$ and $r_3$ are applied.
 %If the configuration is $\{s_2\}$, only $r_3$ is applied. The configurations $\{s_1,s_3\}$,and $\{s_3\}$ make no reaction applicable.\\
 Now, we show the first three steps of an interacting process $\pi=(\gamma,\delta)$ where the context provides entities as follows: $\gamma = \{C_0 = \{s_1,s_2\}, C_1 =\{s_3\},C_2=\{s_1,s_4\},C_3= \emptyset, C_4,\dots, C_n\}$. Initally $D_0 = \emptyset$ and from
 $C_0\cup D_0=\{s_1,s_2\}$ we get to $C_1\cup D_1=\{s_1,s_2,s_3\}$ by applying  $a_1$, $a_2$ (they both are enabled); after that  we get to $C_2\cup D_2=\{s_1,s_2,s_4\}$, by applying $a_2$ only ($a_1$ is not enabled); finally we get to $C_3\cup D_3=\{s_1\}$ by applying $a_1$ ($a_2$ is not enabled). 
Thus $\delta=\{D_0 = \emptyset, D_1=\{s_1,s_2\},D_2=\{s_2\},D_3=\{s_1\},D_4,\dots,D_n\}$.
We remark that at every state $C_i \cup D_i$ all the reactions which are enabled are applied, so the computation is deterministic.
\end{example}
%\end{color}
